Towards a framework for reasoning about stateful compiler IRs using coinductive reasoning.

### 2024-06-25 paper.tex
put references before the dot spelling and grammar pass comment out iris

@@

@@

Formal verification is crucial in ensuring the reliability and
correctness of software systems. In the context of compiler design, this
importance is magnified as compilers are responsible for translating
high-level code into machine-executable instructions. Reasoning about
compiler optimizations is notoriously difficult, and even minor bugs in
widely used compilers can lead to subtle, hard-to-debug miscompilations
with far-reaching consequences. This challenge makes the formal
verification of compiler intermediate representations (IRs) an
especially compelling topic, as it promises to enhance the dependability
of the entire software development lifecycle.

Now, formal verification has been done before. Most notable is the
CompCert project: a fully verified end-to-end C compiler.~\cite{leroyCompCertFormallyVerified} Yet, the vast majority of
code today is still being compiled with non-verified compilers. We ask,
thus, how do we formally verify not \emph{a} compiler, but, how do we
scale formal verification to \emph{mainstream} compilers.

Take, for example, LLVM, a widely-used open-source compiler that is used
to compile, e.g., C, C++, Fortran, Rust, and Swift. Verifying the entire
LLVM codebase, comprising of a few million lines of C++ code, is
infeasible. Instead, we follow the same approach as Lopez et al and
focus on verifying that individual rewrites on a compiler IR preserve
semantics.~\cite{lopesAlive2BoundedTranslation2021}

@@

In the next two years, I'd like to look into a feature of MLIR called
\emph{Graph Regions}. Specific dialects may use such regions to allow a
variable to be used \emph{before} it is defined --- hence, turning the
def-use chain into a general graph. Generally, the semantics of such
dialects are statefull, and cycles are broken by looking at the value of
a variable at a previous state. For example, graph regions are used by
dialects that model hardware,
where variables model wires and cycles represent physical cycles in the
circuit.~\cite{eldridgeMLIRHardwareCompiler}

@@

CompCert is the
seminal work when it comes to verified compilation: an end-to-end
verified to not miscompile compiler for the C language.~\cite{leroyCompCertFormallyVerified}
However, exactly
because it's an end-to-end verified monolith, CompCert is not
competitive compared to traditional non-verified compilers. Outside of
certain high-risk domains, where traditionally optimizations would be
turned off entirely,~\cite{kastnerCompCertPracticalExperience} the
vast majority of C code is thus still compiled with non-verified
compilers, which studies have shown to be prone to miscompilations.~\cite{yangFindingUnderstandingBugs2011}

Lopes et al.~propose \emph{translation validation} as a way to scale
formal verification.~\cite{lopesAlive2BoundedTranslation2021}
Their implementation, called Alive, has found wide adoption in the
development of LLVM. Given a declarative description of an optimization
as a source and target pattern in LLVM IR, the tool uses an SMT solver
to automatically determine whether the optimized pattern exhibits only
behaviours of the original program.

@@

Our group has been developing
LeanMLIR, a framework for giving
semantics to MLIR-style compiler IRs in the Lean theorem prover.~\cite{demouraLeanTheoremProver2015} As a theorem prover, Lean is
much more expressible than SMT solvers. Among theorem provers, Lean is
particularly suited for this purpose because of it's community of
mathematicians developing a comprehensive mathematical library,~\cite{themathlibcommunityLeanMathematicalLibrary2020} 
and because of it's highly extensive syntax and meta-programming capabilities~\cite{ullrichNotationsHygienicMacro2022, paulinoMetaprogrammingLean}.

@@

Lean, like most theorem provers, is built with around a native concept
of inductive types. This is no surprise, given that most mathematical
constructions are inductive in nature. Coinductive constructions, in
contrast, are less common in general mathematics, but, they are crucial
in computer science.

For us, the main difference of interest is that inductive datastructures
are finite, while coinductive datastructures capture infinite
behaviours.~\cite{sangiorgiIntroductionBisimulationCoinduction2011}
In particular, since Lean requires us to prove that recursive functions are
terminating, we need corecursion to correctly model things like infinite
loops and other non-terminating behaviours that are common in compiler
IRs.

For example, Xia et al.~used interaction trees, a coinductive type, in
their Coq formalization of recursive and impure programs.~\cite{xiaInteractionTreesRepresenting2020} Similarly, MLIR has an
``fsm'' dialect~\footnote{https://circt.llvm.org/docs/Dialects/FSM/}
centred on finite state machines. Such automata are understood to be
coinductive in nature, so mechanizing the semanantics of this dialect in
Lean will likely require coinduction.

Consequently, we need a user-friendly way to define coinductive
structures and to formalize coinductive proof methods in Lean. Key prior
work in this area include my own master's thesis (\cite{keizerImplementingDefinitionalCo}), 
which is based on earlier
work by Avigad et al that models coinductive as quotients of polynomial
functors (QPFs).~\cite{avigadDataTypesQuotients2019}

@@

Isabelle, in particular, is relevant because it's construction of
coinductive types, in terms of \emph{bounded natural functors} is
closely related to our construction in terms of QPFs.~\cite{furerQuotientsBoundedNatural2022} 
There are still new
challenges, however, because Lean's dependent types mean we have to
worry about the distinction between definitional equality (which is a
decidable, but weaker notion of equality) and propositional equality
(which captures more things that are intuitively equal, but requires
proof). Isabelle, in contrast, uses a weaker logical system where all
equalities are decidable.

In contrast, the coinduction-in-the-kernel approach allows us to modify
the kernel to understand coinduction and ensure the kernel recognizes
the desired definitonal equalities. However, modifying the trusted
kernel carries a large burden of proof, since the changes could
compromise the logical soundness of the entire system. In fact, the
original implementation of coinduction, called \emph{positive}
coinduction, is nowadays discouraged in favour of an alternative
implementation, since positive coinduction breaks the subject reduction
property\footnote{The subject reduction property states that reducing a program does not change its type}.~\cite{sozeauCorrectCompleteType}
Our coinductives-as-a-library approach requires no new axioms nor
changes to the kernel, and is thus guaranteed not to change any meta
properties of Lean.

@@

The bulk of my time has been spent on developing, in collaboration with
other members of our research group, the LeanMLIR framework, culminating
in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
Compiler IRs''}. As part of this development, we

@@

\subsection{Stretch Goal: Iris}\label{stretch-goal-iris}

\begin{quote}
{[}!TODO{]} Talk about how denotational semantics are cool, and how the
aforementioned improvements will let us define dialects with
side-effects, but reasoning about them will still be a bit painful. We'd
really like to reason with program logics instead, in particular, iris:
\cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
\end{quote}

>>>>

% \usepackage[numbers]{natbib} % for \citeauthor

@@

\newcommand*{\etal}{~\emph{et al.}}

@@

Formal verification is crucial in ensuring the reliability and
correctness of software systems. 
Reasoning about compiler optimizations, in particular, is notoriously difficult, 
and even minor bugs in widely-used compilers can lead to subtle, hard-to-debug miscompilations
with far-reaching consequences. 
This challenge makes the formal verification of compiler intermediate representations (IRs) a
particularly compelling topic, 
as it promises to enhance the dependability of the entire software development lifecycle.

Now, formal verification has been done before. Most notable is the
CompCert project: a fully verified end-to-end C compiler~\cite{leroyCompCertFormallyVerified}. Yet, the vast majority of
code today is still being compiled with non-verified compilers. We ask,
thus, how do we formally verify not just \emph{a} compiler, but, how do we
scale formal verification to \emph{mainstream} compilers.

Take, for example, LLVM, a widely-used open-source compiler that is used
to compile, e.g., C, C++, Fortran, Rust, and Swift. Verifying the entire
LLVM codebase, comprising of a few million lines of C++ code, is
infeasible. Instead, we follow the same approach as Lopes\etal{} by
focussing on verifying that individual rewrites on a compiler IR preserve
semantics~\cite{lopesAlive2BoundedTranslation2021}.

@@

In the next two years, I'd like to look into a feature of MLIR called
\emph{Graph Regions}. Specific dialects may use such regions to allow a
variable to be used \emph{before} it is defined --- hence, turning the
def-use chain into a general graph. Generally, the semantics of such
dialects are statefull, and cycles are broken by looking at the value of
a variable at a previous state. For example, graph regions are used by
dialects that model hardware,
where variables model wires and cycles represent physical cycles in the
circuit~\cite{eldridgeMLIRHardwareCompiler}.

@@

CompCert is the
seminal work when it comes to verified compilation: an end-to-end
verified to not miscompile compiler for the C language~\cite{leroyCompCertFormallyVerified}.
However, exactly
because it's an end-to-end verified monolith, CompCert is not
competitive compared to traditional non-verified compilers. Outside of
certain high-risk domains, where traditionally optimizations would be
turned off entirely,~\cite{kastnerCompCertPracticalExperience} the
vast majority of C code is thus still compiled with non-verified
compilers, which studies have shown to be prone to miscompilations~\cite{yangFindingUnderstandingBugs2011}.

Lopes et al.~propose \emph{translation validation} as a way to scale
formal verification~\cite{lopesAlive2BoundedTranslation2021}.
Their implementation, called Alive, has found wide adoption in the
development of LLVM. Given a declarative description of an optimization
as a source and target pattern in LLVM IR, the tool uses an SMT solver
to automatically determine whether the optimized pattern exhibits only
behaviours of the original program.

@@

Our group has been developing
LeanMLIR, a framework for giving
semantics to MLIR-style compiler IRs in the Lean theorem prover~\cite{demouraLeanTheoremProver2015}. 
As a theorem prover, Lean is
much more expressible than SMT solvers. Among theorem provers, Lean is
particularly suited for this purpose because of it's community of
mathematicians developing a comprehensive mathematical library,~\cite{themathlibcommunityLeanMathematicalLibrary2020} 
and because of its highly extensive syntax and meta-programming capabilities~\cite{ullrichNotationsHygienicMacro2022, paulinoMetaprogrammingLean}.

@@

Lean, like most theorem provers, is built around a native concept
of inductive types. This is no surprise, given that most mathematical
constructions are inductive in nature. Coinductive constructions, in
contrast, are less common in general mathematics, but, they are crucial
in computer science.

For us, the main difference of interest is that inductive datastructures
are finite, while coinductive datastructures capture infinite
behaviours~\cite{sangiorgiIntroductionBisimulationCoinduction2011}.
In particular, since Lean requires us to prove that recursive functions are
terminating, we need corecursion to correctly model things like infinite
loops and other non-terminating behaviours that are common in compiler
IRs.

For example, Xia et al.~used interaction trees, a coinductive type, in
their Coq formalization of recursive and impure programs~\cite{xiaInteractionTreesRepresenting2020}.
Similarly, MLIR has an
``fsm'' dialect~\footnote{https://circt.llvm.org/docs/Dialects/FSM/}
centred on finite state machines. Such automata are understood to be
coinductive in nature, so mechanizing the semanantics of this dialect in
Lean will likely require coinduction.

Consequently, we need a user-friendly way to define coinductive
structures and to formalize coinductive proof methods in Lean. Key prior
work in this area includes my own master's thesis (\cite{keizerImplementingDefinitionalCo}), 
which is based on earlier
work by Avigad et al that models coinductive as quotients of polynomial
functors (QPFs)~\cite{avigadDataTypesQuotients2019}.

@@

Isabelle, in particular, is relevant because it's construction of
coinductive types, in terms of \emph{bounded natural functors} is
closely related to our construction in terms of QPFs~\cite{furerQuotientsBoundedNatural2022}. 
There are still new
challenges, however, because Lean's dependent types mean we have to
worry about the distinction between definitional equality (which is a
decidable, but weaker notion of equality) and propositional equality
(which captures more things that are intuitively equal, but requires
proof). Isabelle, in contrast, uses a weaker logical system where all
equalities are decidable.

In contrast, the coinduction-in-the-kernel approach allows us to modify
the kernel to understand coinduction and ensure the kernel recognizes
the desired definitonal equalities. However, modifying the trusted
kernel carries a large burden of proof, since the changes could
compromise the logical soundness of the entire system. In fact, the
original implementation of coinduction, called \emph{positive}
coinduction, is nowadays discouraged in favour of an alternative
implementation, since positive coinduction breaks the subject reduction
property\footnote{The subject reduction property states that reducing a program does not change its type}~\cite{sozeauCorrectCompleteType}.
Our coinductives-as-a-library approach requires no new axioms nor
changes to the kernel, and is thus guaranteed not to change any meta
properties of Lean.

@@

The bulk of my time has been spent developing the LeanMLIR framework, 
in collaboration with other members of our research group, 
culminating in our ITP 2024 paper \emph{``Verifying Peephole Rewriting In SSA
Compiler IRs''}.

@@

% \subsection{Stretch Goal: Iris}\label{stretch-goal-iris}

% \begin{quote}
% {[}!TODO{]} Talk about how denotational semantics are cool, and how the
% aforementioned improvements will let us define dialects with
% side-effects, but reasoning about them will still be a bit painful. We'd
% really like to reason with program logics instead, in particular, iris:
% \cite{jungIrisMonoidsInvariants2015}\cite{jungHigherorderGhostState2016}
% \end{quote}
